Rates of true polar wander in convecting planets

### 2015-09-08 tpw_rate.tex
Address comments, cleanup

These questions suggest that an approach rooted in dimensional analysis and fluid dynamics can clarify the rates and magnitudes of TPW.
Most previous studies coupling mantle convection models to polar wander calculations have done so with prescribed density perturbations \citep[e.g.][]{greff2004upwelling}, or prescribed moment if inertia variations \citep[e.g.][]{tsai2007theoretical, creveling2012mechanisms}. 
\citet{richards1999polar} coupled thermal convection models to a polar wander model, but did not address in detail the scaling relationships between the two.

@@

\subsection{Rotational deformation}

@@

Splitting the volume into $V_H$ and $\Delta V$ and substituting \eqref{eq:eos} into \eqref{eq:quasistatic_angular_momentum}, we find
\begin{equation}
\begin{aligned}
&\int_{V_H} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV + 
\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV +  \\
&\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV + 
\int_{\Delta V} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV = 0  \\
\end{aligned}
\end{equation}
The first term of this equation is zero due to the hydrostatic equation.  
The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\Gamma$.
Removing these, we find
\begin{equation}
\begin{aligned}
\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV = 
-\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV
\end{aligned}
\end{equation}
This equation may be identified with \eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of intertia with the spin axis (right side).
Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table \ref{tab:nondim}.

\ifdetail
\subsection{Buckingham Pi theorem and nondimensionalization}
The Buckingham Pi theorem (essentially an application of the rank-nullity theorem) allows us to determine the number of nondimensional numbers for this problem.
We count the number of fundamental units for the problem, in this case mass (kg), length (m), time (s), and temperature (K), as well as the number of parameters in use for this problem.  
These parameters are listed in Table \ref{tab:parameters}, and subtracting the ...

@@

If we plug these nondimensionalizations into the governing equations we find that the incompressibility constraint remains unchanged.  
The temperature equation and equation of state become

\begin{equation}
\frac{\partial T}{\partial t} + {\bf v} \cdot \nabla T = \nabla^2 T
\end{equation}

\begin{equation}
\rho = 1 - \alpha \Delta T \;\;  T
\end{equation}

We may furthermore define a hydrostatic reference state where the density is the reference density everywhere and the velocity is zero, i.e.

\begin{equation}
 \rho_0{\bf \Omega \times \Omega \times r}= - \nabla P_0 + \rho_0 {\bf g}
\end{equation}

Subtracting this from the momentum equation and defining a dynamic pressure $P^* = P - P_0$, we find

@@

It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces, or Froude number \citep{munk1960rotation}.
\begin{equation}
(C-A) \sim I_0 \mathrm{Fr}
\end{equation}

@@

\subsection{An estimate of $\Lambda_{ij}$}

@@

At low Rayleigh number we expect the spectrum of the temperature field to be dominated by only a few low-degree modes which are largely influenced by the aspect ratio.
This spectrum may or may not have a lot of power in degree-two, and does not depend strongly on time.

At high Rayleigh number we expect the shortest lengthscales to be limited by the effects of thermal diffusion,
which tends to wipe out thermal heterogeneity at small scales.  Consequently, there will be no power in modes
with shorter lenghtscales than that allowed by diffusion. 
Therefore we expect, to a good approximation, that the infinite sum in Equation \eqref{eq:T_series} can be truncated at some maximum wavenumber, set by the smallest lengthscale $d$:
\begin{equation}
n_{\mathrm{max}}, l_{\mathrm{max}}, m_{\mathrm{max}} \sim \frac{R}{d}
\end{equation}
The total number of modes that are accessible to the system are therefore

@@

We performed a series of numerical simulations of mantle convection at different Rayleigh numbers to test this scaling.
We used the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}, based on the finite element library \texttt{deal.II} \citep{dealII82}.
which allows for flexible implementation of 
In order to test a wide range of Rayleigh numbers, we ran the simulations in a 2D annulus, tracking the eigenvalues of the 
moment of inertia tensor and integrating Equation~\eqref{eq:liouville_ode} in time.
For the 2D simulations there is a reduced dimensionality when calculating the number of modes,
so $N_\mathrm{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\mathrm{degree\textendash two}} \sim \mathrm{Ra}^{-2/3}$, which is used in the simulations shown in Figure~\ref{fig:eigengap}.
This result has a simple interpretation.
As the Rayleigh number of the system increases, the smallest lengthscale of convective features gets smaller.
The total power in the temperature field is spread across a larger spectrum, leaving less total power for the degree-two part, which is what drives TPW.

@@

\subsection{An estimate of the mismatch $\theta$}

@@

The preceding results clarify the complex relationship between the Rayleigh number, the Froude number and the rates of TPW for a convecting planet.
As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
There is a constant competition between growth of the mismatch angle $\theta$ through Equation \eqref{eq:grow_perturbation} and its relaxation through Equation \eqref{eq:simple_milankovitch}.
We may plug in the estimate for $\Lambda_{ij}$ into equation \eqref{eq:simple_milankovitch} to find the strikingly simple expression
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{\mathrm{Fr}} \sin{2 \theta}
\label{eq:simplest_milankovitch}
\end{equation}
Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
At high Rayleigh numbers response timescales for relaxation of the mismatch angle go down.
At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \sim \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{\mathrm{Fr}} \sin{2 \theta}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for $d(\mathrm{Ra})$ will still show a weak dependence on the Rayleigh number.

@@

Indeed, we can identify two endmember behaviors of Equation \eqref{eq:simplest_milankovitch}.
When convection is not sufficiently chaotic to create a large $\theta$ we are in the regime where the planet's rotation axis closely tracks that of the convective moment.
This is the regime considered in \citet{steinberger1997changes}, \citet{roberts2007cause}, and \citet{zhong2007supercontinent}, and can be considered the ``slow TPW'' regime.
When convection is more chaotic, however, there may be large excursions in $\theta$, which can drastically increase the wander rate.  
If $\theta=90^\circ$, this corresponds to IITPW \citep{kirschvink1997evidence}.  
This, however, is a special case in the large $\theta$, ``fast TPW'' regime.

>>>>

These questions suggest that an approach rooted in dimensional analysis and fluid dynamics can clarify the rates and magnitudes of TPW.
Most previous studies coupling mantle convection models to polar wander calculations have done so with prescribed density perturbations \citep[e.g.][]{greff2004upwelling}, or prescribed moment of inertia variations \citep[e.g.][]{tsai2007theoretical, creveling2012mechanisms}. 
\citet{richards1999polar} coupled thermal convection models to a polar wander model, but did not address in detail the scaling relationships between the two.

@@

\subsection{Rotational deformation}
\label{sec:rotational_deformation}

@@

Splitting the volume into $V_H$ and $\Delta V$ and substituting \eqref{eq:eos} into \eqref{eq:quasistatic_angular_momentum}, we find
\begin{equation}
\begin{aligned}
&\int_{V_H} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV + 
\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV +  \\
&\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV + 
\int_{\Delta V} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV = 0  \\
\end{aligned}
\end{equation}
The first term of this equation is zero due to the hydrostatic equation.  
The fourth term is negligible due to being second order in the smallness parameters $\Delta V/V_H$ and $\Gamma\equiv \alpha \Delta T$.
Removing these, we find
\begin{equation}
\begin{aligned}
\int_{\Delta V} \rho_0 {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV = 
-\int_{V_H} \rho_0 \alpha (T-T_0) {\bf r \times \mitbf{\Omega} \times \mitbf{\Omega} \times {\bf r }} \,dV
\end{aligned}
\end{equation}
This equation may be identified with \eqref{eq:disequilibrium}, where disequilibrium in the rotational deformation (left side) is balanced by the mismatch of the convective moment of intertia with the spin axis (right side).
Our goal is to identify characteristic sizes of these quantities, which must be functions of the nondimensional numbers identified in Table \ref{tab:nondim}.

\ifdetail
\subsection{Buckingham Pi theorem and nondimensionalization}
\label{sec:buckingham}
The Buckingham Pi theorem (essentially an application of the rank-nullity theorem) allows us to determine the number of nondimensional numbers for this problem.
We count the number of fundamental units for the problem, in this case mass (kg), length (m), time (s), and temperature (K), as well as the number of parameters in use for this problem.  
These parameters are listed in Table \ref{tab:parameters}, and subtracting the ...

@@

If we plug these nondimensionalizations into the governing equations we find that the incompressibility constraint remains unchanged.  
The temperature equation and equation of state become
\begin{equation}
\frac{\partial T}{\partial t} + {\bf v} \cdot \nabla T = \nabla^2 T
\end{equation}

\begin{equation}
\rho = 1 - \alpha \Delta T \;\;  T
\end{equation}
We may furthermore define a hydrostatic reference state where the density is the reference density everywhere and the velocity is zero, i.e.

\begin{equation}
 \rho_0{\bf \Omega \times \Omega \times r}= - \nabla P_0 + \rho_0 {\bf g}
\end{equation}
Subtracting this from the momentum equation and defining a dynamic pressure $P^* = P - P_0$, we find

@@

It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces, or Froude number \citep{munk1960rotation}:
\begin{equation}
(C-A) \sim I_0 \mathrm{Fr}
\end{equation}

@@

\subsection{An estimate of $\Lambda_{ij}$}
\label{sec:lambda}

@@

At low Rayleigh number we expect the spectrum of the temperature field to be dominated by only a few low-degree modes which are largely influenced by the aspect ratio.
This spectrum may or may not have a lot of power in the degree-two modes, and does not depend strongly on time.

At high Rayleigh number we expect the shortest lengthscales to be limited by the effects of thermal diffusion,
which tends to wipe out thermal heterogeneity at small scales.  Consequently, there will be no power in modes
with shorter lengthscales than that allowed by diffusion. 
Therefore we expect, to a good approximation, that the infinite sum in Equation \eqref{eq:T_series} can be truncated at some maximum wavenumber, set by the smallest lengthscale $d$:
\begin{equation}
n_{\mathrm{max}}, l_{\mathrm{max}}, m_{\mathrm{max}} \sim \frac{R}{d}
\end{equation}
The total number of modes that are accessible to the system are therefore

@@

We performed a series of numerical simulations of mantle convection at different Rayleigh numbers to test this scaling.
We used the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}, based on the finite element library \texttt{deal.II} \citep{dealII82}.
which allows for flexible implementation of different rheologies, geometries, and postprocessors.
In order to test a wide range of Rayleigh numbers, we ran the simulations in a 2D annulus, tracking the eigenvalues of the 
moment of inertia tensor and integrating Equation~\eqref{eq:liouville_ode} in time.
For the 2D simulations there is a reduced dimensionality when calculating the number of modes,
so $N_\mathrm{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\mathrm{degree\textendash two}} \sim \mathrm{Ra}^{-2/3}$, which is used in the simulations shown in Figure~\ref{fig:eigengap}.
This result has a simple interpretation.
As the Rayleigh number of the system increases, the smallest lengthscale of convective features gets smaller.
The total power in the temperature field is spread across a larger spectrum, leaving less total power for the degree-two part, which is what drives TPW.

@@

\subsection{An estimate of the mismatch $\theta$}
\label{sec:theta}

@@

The preceding results clarify the complex relationship between the Rayleigh number, the Froude number and the rates of TPW for a convecting planet.
As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
There is a constant competition between growth of the mismatch angle $\theta$ through Equation \eqref{eq:grow_perturbation} and its relaxation through Equation \eqref{eq:simple_milankovitch}.
We may plug in the estimate for $\Lambda_{ij}$ into equation \eqref{eq:simple_milankovitch} to find the strikingly simple expression
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{\mathrm{Fr}} \sin{2 \theta}
\label{eq:simplest_milankovitch}
\end{equation}
Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
Response timescales for relaxation of the mismatch angle go down at high Rayleigh numbers.
At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \sim \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{\mathrm{Fr}} \sin{2 \theta}
\label{eq:simplest_milankovitch_defer_scaling}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for $d(\mathrm{Ra})$ will still show a weak dependence on the Rayleigh number.

@@

Indeed, we can identify two endmember behaviors of Equation \eqref{eq:simplest_milankovitch_defer_scaling}.
When convection is not sufficiently chaotic to create a large $\theta$ we are in the regime where the planet's rotation axis closely tracks that of the convective moment.
This is the regime considered in \citet{steinberger1997changes}, \citet{roberts2007cause}, and \citet{zhong2007supercontinent}, and can be considered the ``slow TPW'' regime.
When convection is more chaotic, however, there may be large excursions in $\theta$, which can drastically increase the wander rate.  
If $\theta=90^\circ$, this corresponds to IITPW \citep{kirschvink1997evidence}.  
This, however, is a special case in the large $\theta$, ``fast TPW'' regime.
